---
title: "Pronghorn Range"
author: "Lauren Puffer"
format: html
editor: visual
---

## Load packages

```{r}
library(here)
library(terra)
library(janitor)
library(CoordinateCleaner)
library(tmap)
library(lubridate)
library(tidyverse)
library(adehabitatHR)
library(sp)
library(sf)
library(tmap)
```

## Load presence data

Presence data of both Pronghorn and Tule elk comes from the Global Biodiversity Information Facility.

```{r}
# read in pronghorn data 
pronghorn <- read.csv(here("Pronghorn", "pronghorn presence.csv"))

# read in tule elk 
tule <- read.csv(here("Tule Elk", "tule elk presence.csv"))
```

Now we need to clean the date column so that the format is consistent using the lubridate package.

```{r}
# PRONGHORN: get date column into year-month-day format
pronghorn_formatted <- pronghorn |>
  mutate(date = dmy(date)) |>
  mutate(date = ymd(date)) 

# TULE: get date column into year-month-day format
tule_formatted <- tule |>
  mutate(date = dmy(date)) |>
  mutate(date = ymd(date))
```

## Plot Pronghorn and Tule Elk over time

Now that we have our datasets in the proper format, we can create graphs of observed Tule elk and Pronghorn per year using 'ggplot'.

```{r}
# PRONGHORN: Get the number of observations per each year 
year_counts <- pronghorn_formatted |>
  mutate(year = year(date)) |>
  count(year, sort = FALSE)

# Plot occurences 
prong_plot <- ggplot(year_counts, aes(x = year, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = seq(min(year_counts$year),
                                  max(year_counts$year),
                                  by = 1)) +
        geom_vline(xintercept = 1998,
             color = "#22A884FF",
             linewidth = 1) +
        geom_vline(xintercept = 1990,
             color = "#22A884FF",
             linewidth = 1) +
  labs(x = "Year",
       y = "Observations",
       title = "Pronghorn Observations per Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

# Save plot
ggsave("pronghorn_observations_by_year.png",
       plot = prong_plot,
       width = 8,
       height = 5,
       dpi = 300,
       bg = "white")

# TULE ELK: Get the number of observations per each year
year_counts_elk <- tule_formatted |>
  mutate(year = year(date)) |>
  count(year, sort = FALSE)

# Plot occurrences 
tule_plot <- ggplot(year_counts_elk, aes(x = year, y = n)) +
  geom_col() +
  scale_x_continuous(breaks = seq(min(year_counts_elk$year),
                                  max(year_counts_elk$year),
                                  by = 1)) +
      geom_vline(xintercept = 2014,
             color = "#22A884FF",
             linewidth = 1) +
        geom_vline(xintercept = 1988,
             color = "#22A884FF",
             linewidth = 1) +
  labs(x = "Year",
       y = "Observations",
       title = "Tule Elk Observations per Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

# Save plot
ggsave("tule_observations_by_year.png",
       plot = tule_plot,
       width = 8,
       height = 5,
       dpi = 300,
       bg = "white")
```

## Create Minimum Convex Polygons

Now, we can create minimum convex polygons for both Pronghorn and Tule elk. This will be used as an approximation of the home range for both species. First we must make the data into SpatialPoints by defining the coordinates and their reference system.

Load data from iNaturalist with unique observer ID.

```{r}
# read in pronghorn data 
pronghorn <- read.csv(here("Pronghorn", "Pronghorn Carrizo.csv"))

# read in tule elk 
tule <- read.csv(here("Tule Elk", "Tule Elk Carrizo.csv"))
```

Format data so that the year and other temporal data is consistent.

```{r}
# PRONGHORN: get date column into year-month-day format
pronghorn_formatted <- pronghorn |>
  mutate(date = mdy_hms(date)) |>
  mutate(date = ymd(date)) 

# TULE: get date column into year-month-day format
tule_formatted <- tule |>
  mutate(date = mdy_hms(date)) |>
  mutate(date = ymd(date))
```

```{r}
# PRONGHORN MCP

# Define coordinates
coordinates(pronghorn_formatted) <- c("longitude", "latitude")

# Assign WGS84 for CRS
crs(pronghorn_formatted) <- "EPSG:4269"


# Get coordinates in meters 
new_crs <- CRS('+init=epsg:5070')
pronghorn_reproj <- spTransform(pronghorn_formatted, new_crs)

# Create a minimum convex polygon using the x and y attributes of species point
prong_poly <- mcp(pronghorn_reproj, percent = 95, 
                  unin = "m",
                  unout = "m2")

# Get the area of the MCP for Pronghorn
prong_area <- prong_poly@data$area 
# 24.12255 before
# 5019601 now in meters^2


# Plot the polygon
plot(prong_poly)




# TULE ELK

# Define coordinates 
coordinates(tule_formatted) <- c("longitude", "latitude")

# Assign WGS84 for CRS
crs(tule_formatted) <- "EPSG:4269"
 
# Get coordinates in meters 
new_crs <- CRS('+init=epsg:5070')
tule_reproj <- spTransform(tule_formatted, new_crs)

# Create a minimum convex polygon using the x and y attributes of species point
tule_poly <- mcp(tule_reproj, percent = 95, 
                  unin = "m",
                  unout = "m2")


# Get the area of the MCP for Tule
tule_area <- tule_poly@data$area 
#  8.152046 before
#  2112476 now in meters62

# Plot the polygon
plot(tule_poly)
```

Now we need to get the overlap of the two polygons.

```{r}
# Create shape files of both
prong_sf <- st_as_sf(prong_poly)
tule_sf <- st_as_sf(tule_poly)

# Get the area of overlap
overlap <- st_intersection(prong_sf, tule_sf)
area_overlap <-st_area(overlap)
overlap_poly <- as(overlap, "Spatial")

# Get percent overlap
pct_overlap <- area_overlap / (prong_area+tule_area)
# 0.2940009 or 29.4%


overlap_area <- st_intersection(tule_sf, prong_sf)
st_area(overlap_area)

# You can then calculate:
percent_overlap_elkside <- st_area(overlap) / st_area(tule_sf)
# 99.25966% of Tule elk homerange overlaps with Pronghorn
print(percent_overlap_elkside)


percent_overlap_prongside <- st_area(overlap) / st_area(prong_sf)
print(percent_overlap_prongside)
# 41.77298% of Pronghorn homerange overlaps with Tule elk
```

Now we will plot these polygons with their overlap.

```{r}
# Convert sp objects to sf if not already
prong_sf <- st_as_sf(prong_poly)
tule_sf <- st_as_sf(tule_poly)
# overlap should already be sf from st_intersection

# Add species column
prong_sf$species <- "Pronghorn MCP"
tule_sf$species <- "Tule Elk MCP"
overlap$species <- "Overlap"

# Ensure all geometry columns have the same name
st_geometry(prong_sf) <- "geometry"
st_geometry(tule_sf) <- "geometry"
st_geometry(overlap) <- "geometry"

# Keep only species column and bind
all_polygons <- rbind(prong_sf[, "species"], 
                      tule_sf[, "species"], 
                      overlap[, "species"])

class(all_polygons)  # Should be "sf"

# Plot
tmap_mode("plot")
```

Bring in the shapefile of Carrizo plane.

```{r}
carrizo_sf <- st_read(here("Carrizo Plane", "3853-s27_2013_s2_polbound_carrizo.shp"))

crs(carrizo_sf) # EPSG 4326

# Reproject 
carrizo_reproj <- st_transform(carrizo_sf, crs = crs(prong_sf))
```

Make map of home range

```{r}
# Make a map of homerange
my_map <- tm_shape(carrizo_reproj)+
  tm_borders()+
tm_shape(all_polygons) +
  tm_polygons(col = "species",
              palette = c("Tule Elk MCP" = "#FDE725FF", 
                          "Pronghorn MCP" = "#22A884FF", 
                          "Overlap" = "#7AD151FF"),
              title = "Home Range") +
  tm_compass(type = "arrow", position = c("right", "top")) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(legend.outside = TRUE,
            main.title = "Tule Elk and Pronghorn home range overlap")

# Then save it
tmap_save(my_map, filename = here("homerange_map.png"), 
          width = 8, height = 6, dpi = 300)
```
